Abstractions and exemplar effects in the German measure noun phrase alternation

### 2017-12-23 Latex/sections/1intro.tex
More editing in Sec. 1

As for the first issue, there has always been an interest in correlating of probabilistic generalisations extracted from corpus data with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers.''
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.
It would be highly surprising if this were possible, and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
While the same approach is by and large applicable to corpus data.

@@

\hl{Turning to the second issue, }
This entails that a variant of prototype theory with features is assumed \citep{Rosch1978}.
Prototype theory is well suited for modeling constructional choices, but it is just one of several similarity-based theories of classification, the most prominent other framework being Exemplar Theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in experimental settings).
Prototype Theory and Exemplar Theory model essentially the same types of effects and differ mainly in whether they assume higher-level abstractions as part of the cognitive representation of categories (prototype theory) or assume that categories emerge through the storage of many exemplars and similarity classification on those exemplars (exemplar theory).
As \cite{Barsalou1990} shows, the two theories model the same types of effects and are informationally equivalent.
Consequently, experiments which favour one theory over the other use procedural behaviour of subjects, for example the speed of category retrieval.
In very early experiments, \cite{PosnerKeele1968} showed, for example, that highly prototypical unseen exemplars were categorised more easily by subjects compared to less prototypical ones, even if these were included in the learning data.
Since corpus data only show artefacts of production events and we have no experimental access to the speakers or writers performance, one should be sceptical whether corpus analysis alone could ever decide which theory of mental representation is more suitable (see also \citealp[22]{Gries2003} and the \citealp[486--487]{Dabrowska2016} quote above).
%The corpus-based approach in \cite{DivjakArppe2013}, for example, primarily shows that corpus data are compatible with aspects of prototype and exemplar theory.

\cite{VanpaemelStorms2008,LeeVanpaemel2008} proposed the \textit{varying abstraction model} (VAM) which ``attempt[s] to balance economy and informativeness'' \citep[745]{LeeVanpaemel2008}, treating models with full abstraction (prototypes) and no abstraction at all (radical exemplar theory) as special cases of a model which allows for both abstraction and exemplar effects.
However, \cite[183--184]{Vanpaemel2016} describes the state of affairs between adherents of neo-prototype theory (such as \citealp{MindaSmith2001,MindaSmith2002}) and exemplar theory as a stalemate, and suggests yet another approach to solve the divide.

\cite{ConawayKurtz2016} show that reference point approaches to category learning such as prototype and exemplar theory fail to explain certain experimental results where subjects learn to generalise beyond the input in a way that cannot be explained by similarity.

\cite{GriffithsEa2009} present a computational model based on the hierarchichal Dirirchlet process which is able to chose the appropriate complexity of representation for a given category.

In \cite{HahnEa2010}, the authors show that subjects even use similarity (exemplar) effects when they are given very simple explicit rules, which is highly relevant because most studies focus on implicit rule-based knowledge, which involves many auxiliary assumptions in actual experiments \citep[2]{HahnEa2010}.

An aspect generally neglected in models of categorisation is how categorisation feeds exemplar generation and creativity (\ie category generation), two subjects of utmost importance to many linguists.
In \cite{JernKemp2013}, a model is assumed where cognitive agents sample from learned distributions over both categories and exemplars to generate new ones.

One thing that should be kept in mind is that most of the research on categorisation in cognitive science uses experiments with highly simplified stimuli and very simple tasks.
It is remarkable in this context that \cite{VoorspoelsEa2011} consider the experimental task reported by them -- assigning typicality scores to nouns from the domains of \textit{animals} and \textit{artefacts} to categories like \textit{bird}, \textit{fish}, \textit{clothing}, or \textit{tools} -- a study of ``superordinate natural language categories, whereas most evidence supporting exemplar representations has been found in artificial categories of a more subordinate level'' \citep[1013]{VoorspoelsEa2011}.
Linguists (not just of the cognitive variety) are usually interested in high-level generalisations and complex categories, especially in (morpho-)syntax.%
\footnote{Notice that recently, approaches have emerged which solve at least some problems by abandoning typically linguistic high-level features altogether \citep{BaayenEa2016,RamscarPort2016}.}
Since, as a consequence, formal models of categorisation in cognitive linguistics have not been spelled out at the same level of mathematical detail as in cognitive science, I propose that the primary focus should be on determining which factors influence choices made by speakers.
As \cite[22]{Gries2003} put it with reference to classic prototype theory, ``even if the form of analysis does not translate into statements on mental representations, the high predictive power [...] shows that the cognitive factors underlying the choice of construction have been identified properly and weighted in accordance with their importance for actual usage.''
A major advancement since this early time has been the focus on which effects are at least more plausibly modelled as prototype\slash abstraction and exemplar effects, and how observed data fits the two approaches (see, for example, \citealp{DivjakArppe2013}).
This paper contributes this discussion.

>>>>

As for the first issue, there has always been an interest in correlating probabilistic generalisations extracted from corpus data with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers.''
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.
It would be highly surprising if this were possible, and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
While the same approach is by and large applicable to corpus data.

@@

\hl{Turning to the second issue, }
This entails that a variant of prototype theory with features is assumed \citep{Rosch1978}.
Prototype theory is well suited for modeling constructional choices, but it is just one of several similarity-based theories of classification, the most prominent other framework being exemplar theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in different experimental settings).
Prototype theory and exemplar theory model essentially the same types of effects but differ significantly in whether they assume higher-level abstractions in the form of single maximally prototypical exemplars or their features (prototype theory) or assume that categories emerge through the storage of many exemplars and similarity classification on those exemplars (exemplar theory).
\hl{Parallel to Langacker here}

As \cite{Barsalou1990} already showed, however, prototype and exemplar theory model the same types of effects and are informationally equivalent.
Consequently, experiments which favour one theory over the other use procedural behaviour of subjects, for example the speed of category retrieval, and not mere output data.
In very early experiments, \cite{PosnerKeele1968} showed, for example, that highly prototypical unseen exemplars were categorised more easily by subjects compared to less prototypical ones, even if these were included in the learning data.
Since corpus data only show artefacts of production events and we have no experimental access to the speaker's or writer's performance, one should be sceptical whether corpus analysis alone could ever decide which theory of mental representation is more suitable (see also \citealp[22]{Gries2003} and the \citealp[486--487]{Dabrowska2016} quote above).

In cognitive science, it is mostly accepted that exemplar theories have greater explanatory power \citep[184]{Vanpaemel2016}, and that abstraction is only needed marginally, if at all.%
\footnote{The hard empirical evidence in favour of exemplar models is substantial.
  For example, in \cite{HahnEa2010}, the authors show that subjects even use exemplar similarity over abstract knowledge even when they are given very simple explicit rules.
  This is highly relevant because most other studies focus on the learning of implicit rule-based knowledge, which involves many auxiliary assumptions in actual experiments \citep[2]{HahnEa2010}.
On the other hand, there is evidence that neither theory is fully adequate to model humans' capabilities to form categories.
For example, \cite{ConawayKurtz2016} show that reference point approaches to category learning such as prototype and exemplar theory fail to explain certain experimental results where subjects learn to generalise beyond the input in a way that cannot be explained by similarity.
}
Still, various attempts have been made over the past decades to settle the dispute between abstraction-based models (models with rules or prototypes) and exemplar models or to find models which unite the two extremes.
\cite{VanpaemelStorms2008,LeeVanpaemel2008} proposed the \textit{varying abstraction model} (VAM) which ``attempt[s] to balance economy and informativeness'' \citep[745]{LeeVanpaemel2008}, treating models with full abstraction (prototypes) and no abstraction at all (radical exemplar theory) as special cases of a model which allows for both abstraction and exemplar effects.
The mixture model of categorisation (MMC) by \cite{Rosseel2002} is a model with abstraction in the form of hierarchical clusters of exemplars, and these clusters of objects are characterised by a probability distribution over their features, and categorising new objects is a process of estimating the probability of this object being from one of the clusters.
\cite{GriffithsEa2009} go further and present a computational model based on the hierarchical Dirirchlet process which is able to chose the appropriate complexity of representation for a given category.
However, despite these (and more) attempts to , \cite[183--184]{Vanpaemel2016} describes the state of affairs between adherents of neo-prototype theory (such as \citealp{MindaSmith2001,MindaSmith2002}) and exemplar theory as a stalemate, and suggests yet another approach to solve the divide.

%An aspect generally neglected in models of categorisation is how categorisation feeds exemplar generation and creativity (\ie category generation), two subjects of utmost importance to many linguists.
%In \cite{JernKemp2013}, a model is assumed where cognitive agents sample from learned distributions over both categories and exemplars to generate new ones.

In cognitive linguistics, \cite{DivjakArppe2013} is a very rare example of a paper where such issues are taken up.
Their corpus-based approach shows ``one way of systematically analyzing usage data as contained in corpora to yield a scheme, compatible with usage-based theories of language, by which the assumptions of both the prototype and exemplar theories can be operationalized'' \citep[267]{DivjakArppe2013}.
Their approach to implementing a varying abstraction model \citep[254--260]{DivjakArppe2013} is based on hierarchical clustering of annotated properties of sentences.
They hierarchically cluster sentences containing Russian verbs of trying.
Then, they single out one sentence from each cluster which scores the highest probability for any of the six \textit{try} verbs according to a polytomous regression model estimated on the same data.
The clusters are interpreted as intermediate-level exemplar-derived abstractions of typical contexts for these high-probability verbs (typically more than one cluster for each verb; \citealp[255--256]{DivjakArppe2013}).
This is pioneering work, but the crucial difference between data-driven corpus-based analyses and experiments in cognitive science (they use \citealp{VerbeemenEa2007} as their reference) is that cognitive experimental research is based on experiments where subjects produce category assignments, and in corpus studies, the categories and category membership is determined purely from the data.

One thing that should be kept in mind is that most of the research on categorisation in cognitive science uses experiments with highly simplified stimuli and very simple tasks.
It is remarkable in this context that \cite{VoorspoelsEa2011} consider the experimental task reported by them -- assigning typicality scores to nouns from the domains of \textit{animals} and \textit{artefacts} to categories like \textit{bird}, \textit{fish}, \textit{clothing}, or \textit{tools} -- a study of ``superordinate natural language categories, whereas most evidence supporting exemplar representations has been found in artificial categories of a more subordinate level'' \citep[1013]{VoorspoelsEa2011}.
Linguists (not just of the cognitive variety) are usually interested in much more complex high-level categories and use large and complex feature sets, especially in (morpho-)syntax.%
\footnote{Notice that recently, approaches have emerged which solve at least some problems by abandoning linguistic high-level features altogether \citep{BaayenEa2016,RamscarPort2016}.}

Most likely a consequence of linguists' interest in  is that formal models of categorisation in cognitive linguistics have not been spelled out at the same level of mathematical detail as in cognitive science.%
\footnote{Again using \cite{DivjakArppe2013} as an example, it is obvious that they use advanced statistics in the form of polytomous regression and hierarchical agglomerative clustering.
However, models like the one proposed in \cite{GriffithsEa2009} are fully spelled out probability models of the behaviour of cognitive agents.}
However, the fact that linguists deal with real data should be seen as an advantage in that it greatly improves the \textit{external validity} of studies, \ie the generalisability of the findings.
Typical artificial tasks in cognitive science have been criticised exactly for the lack of external validity (\eg \citealp{Murphy2003}).

I propose that the primary focus should be on determining which factors influence choices made by speakers.
As \cite[22]{Gries2003} put it with reference to classic prototype theory, ``even if the form of analysis does not translate into statements on mental representations, the high predictive power [...] shows that the cognitive factors underlying the choice of construction have been identified properly and weighted in accordance with their importance for actual usage.''
A major advancement since this early time has been the focus on which effects are at least more plausibly modelled as prototype\slash abstraction and exemplar effects, and how observed data fits the two approaches (see, for example, \citealp{DivjakArppe2013}).
This paper contributes this discussion.
